### 2024-06-04 thesis/input_files/paper1.tex
cleaned all problems and warnings

We take then the inner product in the high dimensional space $\mathcal{H}$ mapped by $\phi(x)$, i.e.\
\\
\begin{align*}
k(x,y):=&\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}
\end{align*}
\\
where $\phi(x)$ is referred to as feature map while $k$ is the kernel function.
\\
Therefore, we can kernelize any algorithm involving a dot product by substituting $\langle x, y \rangle_{\mathcal{X}}$ with $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$
\\
One would expect constructing the feature maps explicitly and then evaluating their inner product in $\mathcal{H}$ to be computationally expensive, and indeed it is. However, we do not have to explicitly perform such calculations. This is because of the kernel trick.
\\
To illustrate the idea behind the kernel trick consider the following example.
Suppose $x, y \in \mathbb{R}^2$ and assume $\phi(x)=(x_{1}^{2},x_{2}^{2},\sqrt{2}x_{1}x_{2})$, then the inner product in the feature space is $x_{1}^{2}y_{1}^{2}+x_{2}^{2}y_{2}^{2}+2x_{1}x_{2}y_{1}y_{2}$.
Notice that this is the same of $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$; thus the kernel trick consists of just using $k(x,y):=(x^\intercal y)^2$.
For instance:
\\
$$
\begin{aligned}
\langle \phi([2,4]),\phi([8,9]) \rangle_{\mathcal{H}}&=  \left[ {\begin{array}{cccc}
    4 & 16 &     \sqrt{2}\cdot 2 \cdot 4\\
  \end{array} } \right]
  \left[ {\begin{array}{cccc}
    64\\
    81\\
    \sqrt{2}\cdot8 \cdot 9
    \\
  \end{array} } \right]\\ &=4\cdot 64 +16\cdot 81 +2 \cdot 2 \cdot 4 \cdot 8 \cdot 9=2704
\end{aligned}
$$
\\
\\
While $k([2,4], [8,9])=(16+36)^2=2704$
We can see that the latter calculation is much quicker and compact.
This gives a taste of how powerful kernels are thanks to the kernel trick.
Indeed, the idea of the the kernel trick can be extended to feature maps
$\phi$ involving an infinite feature space. In these cases, calculating the
respective kernel is equal to calculating the inner product between an infinite
number of features of the data points.

@@

Next, we make this characterisation explicit by showing how to construct a feature mapping $\phi$ of a kernel $k$ such that it is positive semidefinite.
Let the feature space be the set of functions 
\\
$$ 
\begin{aligned}
    \mathcal{F} =\bigg\{ \sum\limits_{i=1}^n \alpha_i k(x_i, \cdot): n \in \mathbb{N}, x_i, y_i \in \mathcal{X}, a_i \in \mathbb{R}, i=1,\cdots, n \bigg\}    
\end{aligned}
$$.
This space is clearly closed under addition of functions and scalar multiplication. Therefore, it is a vector space.
Letting $f(x)=\sum\limits_{i=1}^n \alpha_i k_i(x_i,x)$ and $g(x)=\sum\limits_{i=1}^m \beta_i k_i(y_i,x)$,  we can introduce an inner product on $\mathcal{F}$ as 
\\
$$
\begin{aligned}
    \langle f,g \rangle_{\mathcal{F}}:=\sum\limits_{i=1}^n\sum\limits_{j=1}^m a_i \beta_j k_i(x_i,y_j)=\sum\limits_{i=1}^n a_i g(x_i)= \sum\limits_{j=1}^m \beta_j f(y_j)
\end{aligned}
$$
\\
Notice, $\langle f, f \rangle_{\mathcal{F}}= \alpha^\intercal K \alpha \geq 0 \ \forall f \in \mathcal{F}$, where $K$ is the kernel matrix.
Moreover, $\langle f, g \rangle_{\mathcal{F}}$ is real valued, symmetric and bilinear and, hence, an inner product.
\\
Additionally, letting $g(x)=k(\cdot, x)$, we have that 
\begin{equation}
    \langle f, k(\cdot, x)\rangle_{\mathcal{F}}=\sum\limits_{i=1}^n a_i k(x_i,x)=f(x)    
\end{equation}
This is known as the reproducing property.
\\
Finally, we extend $\mathcal{F}$ to a complete vector space.
Consider a fixed input $x$ and a Cauchy sequence $\{f_n\}_{n=1}^{\infty}$, we then have by means of the Cauchy-Schwarz inequality that 
\\
\begin{equation}
    (f_n(x)- f_m(x))^2=\langle f_n-f_m, k(\cdot, x)\rangle_{\mathcal{F}}^2 \leq \| f_n - f_m \|_{\mathcal{F}}^2 k(x,x)    
\end{equation}
\\
We have then, that $f_n(x)$ is a Cauchy sequence of real numbers. Thus, we can let $g(x):=\underset{n\to \infty}\lim f_n(x)$ and add all such limits functions to $\mathcal{F}$, doing so we complete $\mathcal{F}$ and we obtain the Hilbert space $\mathcal{H}$ associated to the kernel $k$.
Having constructed the feature space $\mathcal{H}$, the mapping $\phi(x)$ is defined as
\\
\begin{equation}
    \phi: x \in \mathcal{X} \to \phi(x)=k(\cdot, x) \in \mathcal{H}
\end{equation}
\\
Notice, that by the reproducing property, $f$ can be represented as the inner product with itself in the feature space, that is 
\begin{equation}
    f(x)=\langle f, \phi(x)\rangle_{\mathcal{H}}=\langle f, k(\cdot, x)\rangle_{\mathcal{H}}=f(x)
\end{equation}

@@

\begin{proof}
    Consider a finite set of points $x_1, \cdots, x_n$, a vector $z \in \mathbb{R}^n$ and let $K_1$ and $K_2$ be the kernel matrices associated to $k_1$ and $k_2$ evaluated on these points.
    \\
    We have:
    \\
    $$  
    \begin{aligned}
    z^\intercal (K_1+K_2) z&= z^\intercal K_1 z + z^\intercal K_2 z \geq 0
    \\
    z^\intercal aK z&= a z^\intercal K z \geq 0
    \end{aligned}
    $$
    Let $K=K_1 \otimes K_2$, we have that the eigenvalues of $K$ are made up by the product pairs of the eigenvalues of $K_1$ and $K_2$. Thus, $K$ is positive semidefinite. Next, the matrix given by $k_1 \cdot k_2$ corresponds to the Schur product of $K_1$ and $K_2$, denote it $H$. Notice, $H$ is a principal submatrix of $K$, hence for any $z \in \mathbb{R}^n$ there exist a $t \in \mathbb{R}^{n^2}$ such that $z^\intercal H z= t^\intercal K t$. Using the fact that K is positive semidefinite we have that for any $z$, it holds 
    $$
    \begin{aligned}
        z^\intercal H z= t^\intercal K t\geq 0
    \end{aligned}
    $$
\end{proof}

\section{Kernel families}\label{kernel families}
The selection of the kernel determines the class of functions that will be searched by the learning algorithm. Prior knowledge of the problem domain can help restricting the candidate families of kernel functions.
Table \ref{tab:kernel types} contains popular kernel families in literature and applications.
\begin{table}
    \caption{Kernel types}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{lll}
        \toprule
       Kernel function & Equation & Hyperparameters \\
       \midrule
       Linear &  $k(x,y)=x^\intercal y$ &   \\
       Polynomial &  $k(x,y)=(x^\intercal y+c)^d$ &   c, d\\
       Gaussian RBF &  $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}^2}{2\l^2}\right)$ &   $l$\\
       Exponential RBF/Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}}{l}\right)$ &   $l$\\
       Absolute Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{1}}{l}\right)$ &   $l$\\
       Hyperbolic/Sigmoid Kernel &  $k(x,y2)=\tanh(\gamma x^\intercal y+r)$ &  $\gamma$, r \\
       Periodic &  $k(x,y)=\exp\left(\frac{-2 sin^2\left(\frac{\pi}{p}\|x-y\|_{2} \right)}{l^2}\right) $ &   $p, l$\\
       Chi-squared kernel &  $k(x,y)=\exp\left(-\gamma \sum\limits_i \frac{(x_i - y_i)^2}{x_i+y_i}\right) $ &   $\gamma$\\
        Cosine  &  $k(x,y)=\frac{x^\intercal y}{\|x\|_{2}\|y\|_{2}}$ &   \\
       Matern  &  $k(x, y) =  \frac{1}{\Gamma(\nu)2^{\nu-1}}\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}
        \Bigg)^\nu Y_\nu\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}\Bigg)$ &  $l, \nu$ \\

@@

Notice, in the Matern family, $\Gamma(\cdot)$ is the gamma function while $Y(\cdot)$ is the Bessel function of the second kind.
\\
When $\nu=p+\frac{1}{2}$, $p \in \mathbb{N}$, the Matern kernel can be rewritten as a product of an exponential and a polynomial of degree $p$. 
\\
$$k(x,y)=\exp \left(-{\frac {{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right){\frac {p!}{(2p)!}}\sum\limits_{i=0}^{p}{\frac {(p+i)!}{i!(p-i)!}}\left({\frac {2{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right)^{p-i}$$
\\
For example:
\begin{itemize}
    \item $p=0,\nu=\frac{1}{2} \implies k(x,y)=\exp\left(-\frac{\|x-y\|_{2}}{l}\right)$
    \item $p=1,\nu=\frac{3}{2} \implies 
    k(x,y)=\left(1+{\frac {{\sqrt {3}}\|x-y\|_{2}}{l }}\right)\exp \left(-{\frac {{\sqrt {3}}\|x-y\|_{2}}{l }}\right)$
    \item $p=2,\nu=\frac{5}{2} \implies k(x,y)=\left(1+{\frac {{\sqrt {5}}\|x-y\|_{2}}{l }}+{\frac {5\|x-y\|_{2}^{2}}{3 l ^{2}}}\right)\exp \left(-{\frac {{\sqrt {5}}\|x-y\|_{2}}{l }}\right)$
\end{itemize}

>>>>

We take then the inner product in the high dimensional space $\mathcal{H}$ mapped by $\phi(x)$, i.e.\

\begin{align*}
k(x,y):=&\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}
\end{align*}

where $\phi(x)$ is referred to as feature map while $k$ is the kernel function.
Therefore, we can kernelize any algorithm involving a dot product by substituting $\langle x, y \rangle_{\mathcal{X}}$ with $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$

One would expect constructing the feature maps explicitly and then evaluating their inner product in $\mathcal{H}$ to be computationally expensive, and indeed it is. However, we do not have to explicitly perform such calculations. This is because of the kernel trick.
To illustrate the idea behind the kernel trick consider the following example.
Suppose $x, y \in \mathbb{R}^2$ and assume $\phi(x)=(x_{1}^{2},x_{2}^{2},\sqrt{2}x_{1}x_{2})$, then the inner product in the feature space is $x_{1}^{2}y_{1}^{2}+x_{2}^{2}y_{2}^{2}+2x_{1}x_{2}y_{1}y_{2}$.
Notice that this is the same of $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$; thus the kernel trick consists of just using $k(x,y):=(x^\intercal y)^2$.
For instance:

$$
\begin{aligned}
\langle \phi([2,4]),\phi([8,9]) \rangle_{\mathcal{H}}&=  \left[ {\begin{array}{cccc}
    4 & 16 &     \sqrt{2}\cdot 2 \cdot 4\\
  \end{array} } \right]
  \left[ {\begin{array}{cccc}
    64\\
    81\\
    \sqrt{2}\cdot8 \cdot 9
    \\
  \end{array} } \right]\\ &=4\cdot 64 +16\cdot 81 +2 \cdot 2 \cdot 4 \cdot 8 \cdot 9=2704
\end{aligned}
$$

While $k([2,4], [8,9])=(16+36)^2=2704$
We can see that the latter calculation is much quicker and compact.
This gives a taste of how powerful kernels are thanks to the kernel trick.
Indeed, the idea of the the kernel trick can be extended to feature maps
$\phi$ involving an infinite feature space. In these cases, calculating the
respective kernel is equal to calculating the inner product between an infinite
number of features of the data points.

@@

Next, we make this characterisation explicit by showing how to construct a feature mapping $\phi$ of a kernel $k$ such that it is positive semidefinite.
Let the feature space be the set of functions

$$ 
\begin{aligned}
    \mathcal{F} =\bigg\{ \sum\limits_{i=1}^n \alpha_i k(x_i, \cdot): n \in \mathbb{N}, x_i, y_i \in \mathcal{X}, a_i \in \mathbb{R}, i=1,\cdots, n \bigg\}    
\end{aligned}
$$
This space is clearly closed under addition of functions and scalar multiplication. Therefore, it is a vector space.
Letting $f(x)=\sum\limits_{i=1}^n \alpha_i k_i(x_i,x)$ and $g(x)=\sum\limits_{i=1}^m \beta_i k_i(y_i,x)$,  we can introduce an inner product on $\mathcal{F}$ as

$$
\begin{aligned}
    \langle f,g \rangle_{\mathcal{F}}:=\sum\limits_{i=1}^n\sum\limits_{j=1}^m a_i \beta_j k_i(x_i,y_j)=\sum\limits_{i=1}^n a_i g(x_i)= \sum\limits_{j=1}^m \beta_j f(y_j)
\end{aligned}
$$

Notice, $\langle f, f \rangle_{\mathcal{F}}= \alpha^\intercal K \alpha \geq 0 \ \forall f \in \mathcal{F}$, where $K$ is the kernel matrix.
Moreover, $\langle f, g \rangle_{\mathcal{F}}$ is real valued, symmetric and bilinear and, hence, an inner product.

Additionally, letting $g(x)=k(\cdot, x)$, we have that 
\begin{equation}
    \langle f, k(\cdot, x)\rangle_{\mathcal{F}}=\sum\limits_{i=1}^n a_i k(x_i,x)=f(x)    
\end{equation}
This is known as the reproducing property.

Finally, we extend $\mathcal{F}$ to a complete vector space.
Consider a fixed input $x$ and a Cauchy sequence $\{f_n\}_{n=1}^{\infty}$, we then have by means of the Cauchy-Schwarz inequality that

\begin{equation}
    (f_n(x)- f_m(x))^2=\langle f_n-f_m, k(\cdot, x)\rangle_{\mathcal{F}}^2 \leq \| f_n - f_m \|_{\mathcal{F}}^2 k(x,x)    
\end{equation}

We have then, that $f_n(x)$ is a Cauchy sequence of real numbers. Thus, we can let $g(x):=\underset{n\to \infty}\lim f_n(x)$ and add all such limits functions to $\mathcal{F}$, doing so we complete $\mathcal{F}$ and we obtain the Hilbert space $\mathcal{H}$ associated to the kernel $k$.
Having constructed the feature space $\mathcal{H}$, the mapping $\phi(x)$ is defined as

\begin{equation}
    \phi: x \in \mathcal{X} \to \phi(x)=k(\cdot, x) \in \mathcal{H}
\end{equation}

Notice, that by the reproducing property, $f$ can be represented as the inner product with itself in the feature space, that is 
\begin{equation}
    f(x)=\langle f, \phi(x)\rangle_{\mathcal{H}}=\langle f, k(\cdot, x)\rangle_{\mathcal{H}}=f(x)
\end{equation}

@@

\begin{proof}
    Consider a finite set of points $x_1, \cdots, x_n$, a vector $z \in \mathbb{R}^n$ and let $K_1$ and $K_2$ be the kernel matrices associated to $k_1$ and $k_2$ evaluated on these points.

    We have:

    $$  
    \begin{aligned}
    z^\intercal (K_1+K_2) z&= z^\intercal K_1 z + z^\intercal K_2 z \geq 0
    \\
    z^\intercal aK z&= a z^\intercal K z \geq 0
    \end{aligned}
    $$
    Let $K=K_1 \otimes K_2$, we have that the eigenvalues of $K$ are made up by the product pairs of the eigenvalues of $K_1$ and $K_2$. Thus, $K$ is positive semidefinite. Next, the matrix given by $k_1 \cdot k_2$ corresponds to the Schur product of $K_1$ and $K_2$, denote it $H$. Notice, $H$ is a principal submatrix of $K$, hence for any $z \in \mathbb{R}^n$ there exist a $t \in \mathbb{R}^{n^2}$ such that $z^\intercal H z= t^\intercal K t$. Using the fact that K is positive semidefinite we have that for any $z$, it holds 
    $$
    \begin{aligned}
        z^\intercal H z= t^\intercal K t\geq 0
    \end{aligned}
    $$
\end{proof}

\section{Kernel families}\label{kernel families}
The selection of the kernel determines the class of functions that will be searched by the learning algorithm. Prior knowledge of the problem domain can help restricting the candidate families of kernel functions.
Table \ref{tab:kernel types} contains popular kernel families in literature and applications.
\begin{table}
    \caption{Kernel types}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{lll}
        \toprule
       Kernel function & Equation & Hyperparameters \\
       \midrule
       Linear &  $k(x,y)=x^\intercal y$ &   \\
       Polynomial &  $k(x,y)=(x^\intercal y+c)^d$ &   c, d\\
       Gaussian RBF &  $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}^2}{2 l^2}\right)$ &   $l$\\
       Exponential RBF/Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{2}}{l}\right)$ &   $l$\\
       Absolute Laplacian& $k(x,y)=\exp\left(-\frac{\|x-y\|_{1}}{l}\right)$ &   $l$\\
       Hyperbolic/Sigmoid Kernel &  $k(x,y2)=\tanh(\gamma x^\intercal y+r)$ &  $\gamma$, r \\
       Periodic &  $k(x,y)=\exp\left(\frac{-2 sin^2\left(\frac{\pi}{p}\|x-y\|_{2} \right)}{l^2}\right) $ &   $p, l$\\
       Chi-squared kernel &  $k(x,y)=\exp\left(-\gamma \sum\limits_i \frac{(x_i - y_i)^2}{x_i+y_i}\right) $ &   $\gamma$\\
        Cosine  &  $k(x,y)=\frac{x^\intercal y}{\|x\|_{2}\|y\|_{2}}$ &   \\
       Matern  &  $k(x, y) =  \frac{1}{\Gamma(\nu)2^{\nu-1}}\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}
        \Bigg)^\nu Y_\nu\Bigg(
        \frac{\sqrt{2\nu}}{l} \|x-y|\|_{2}\Bigg)$ &  $l, \nu$ \\

@@

Notice, in the Matern family, $\Gamma(\cdot)$ is the gamma function while $Y(\cdot)$ is the Bessel function of the second kind.
When $\nu=p+\frac{1}{2}$, $p \in \mathbb{N}$, the Matern kernel can be rewritten as a product of an exponential and a polynomial of degree $p$.

\begin{equation*}
    \begin{split}
    k(x,y)&=\exp \left(-{\frac {{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right){\frac {p!}{(2p)!}}\\
    & \ \cdot \sum\limits_{i=0}^{p}{\frac {(p+i)!}{i!(p-i)!}}\left({\frac {2{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right)^{p-i}
\end{split}
\end{equation*}

For example:
\begin{itemize}
    \item $p=0,\nu=\frac{1}{2} \implies k(x,y)=\exp\left(-\frac{\|x-y\|_{2}}{l}\right)$
    \item $p=1,\nu=\frac{3}{2} \implies 
    k(x,y)=\left(1+{\frac {{\sqrt {3}}\|x-y\|_{2}}{l }}\right)\exp \left(-{\frac {{\sqrt {3}}\|x-y\|_{2}}{l }}\right)$
    \item $p=2,\nu=\frac{5}{2} \implies k(x,y)=\left(1+{\frac {{\sqrt {5}}\|x-y\|_{2}}{l }}+{\frac {5\|x-y\|_{2}^{2}}{3 l ^{2}}}\right)\exp \left(-{\frac {{\sqrt {5}}\|x-y\|_{2}}{l }}\right)$
\end{itemize}
